---
title: "Group Work 1 - Predictive modeling"
author: "Donjet Dzemaili, Fereshteh Ahmadi, Michael Etter, Seraina Zimmermann"
date: "2025-05-26"
output: html_document
---

# Introduction
What is the assignment all about. 

# Load and prepare the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(writexl)
library(readxl)

# Load the data
load("data_wage.RData")

# Rename for clarity
data_original <- data
rm(data)  # Optional: remove old name to avoid confusion

```

Make a copy from the original dataset to excel. so all of the group members can answer the questions and fill them directly in a separated row in the excel file. 
```{r export csv file to data}

# Export to Excel
write_xlsx(data_original, "data_wage_editable.xlsx")

```

After everyone filled the excel file out, we load the filled excel file. 

```{r upload filled excel file}

# Read the filled-in Excel file
data <- read_excel("data_wage_filled.xlsx")

```

```{r }

names(data)
str(data)

```

# Preprocessing

```{r export csv file to data}

names(data)
str(data)

```


# Training and Testing

```{r export csv file to data}

names(data)
str(data)

```

# Explanation of the model

```{r export csv file to data}

names(data)
str(data)

```

# Prediction of our wages

```{r export csv file to data}

names(data)
str(data)

```

