---
title: "Group Work 1 - Predictive modeling"
author: "Donjet Dzemaili, Fereshteh Ahmadi, Michael Etter, Seraina Zimmermann"
date: "2025-05-26"
output: html_document
---

HALLO DONJET - UPLOADE DS FILE JITZE. ZEIT: 21:52 - LÖSCH DIE LINE USE - ISCH NUR ZUR INFO

# Introduction
What is the assignment all about. Donjet is the King! - Yessir

# Load and prepare the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(writexl)
library(readxl)
library(knitr)


# Load the data
load("data_wage.RData")

# Rename 
data_original <- data
rm(data)  # Remove old name to avoid confusion

```

Make a copy from the original dataset to excel. so all of the group members can answer the questions and fill them directly in a separated row in the excel file. 
```{r export csv file to data}

# Export to Excel
write_xlsx(data_original, "data_wage_editable.xlsx")

```

After everyone filled the excel file out, we load the filled excel file. 

```{r upload filled excel file}

# Read the filled-in Excel file
data <- read_excel("data_wage_filled.xlsx")

```

# Data Exploration
Explore the data

## Dimensions of the dataset

```{r dimensions }
dim(data)

```
## First and last few rows

```{r head}

# Display the first few rows
kable(
  head(data),
  caption = "First Few Rows of the Dataset"
)

```

```{r tail}

# Display the last few rows
kable(
  tail(data),
  caption = "Last Few Rows of the Dataset"
)

```

## Structure of the dataset
maybe also display a summary of how many numerical, categorical, ... variables there are in the dataset (different data types)

```{r structure }

str(data)

```

# Feature Selection
Clear justification for selected features. why did we chose the features we have chosen

Use this to determine: 
t-test for numerical variables?
calculate p-value for numerical variables?
Or Boruta algorithm?

# Data Quality Issues 

## Completeness
missing values
na's

```{r completeness }

str(data)

```

## Duplicates

```{r duplicates }

str(data)

```

## Data Consistency
Convert categorical variables to factors (Explain why)
(Optionally) one-hot encode factors

```{r data consistency }

str(data)

```

## Outliers

```{r completeness }

str(data)

```

## Standardization

```{r standardization}


```

# Model choice
Describe the chosen model
Why did we chose this model (proper explanation)

# Training
explain how we train and why we train the way we train

```{r training}



```

# Testing
explain the used evaluation metrics, justify - why do we use the metrics we use

Then evaluate performance of the model

```{r testing}



```

# Explanation of the model
Describe the used explanation techniques
Run explainability techniques
Clear understanding of which features drive predictions and how they influence wage predictions. 

(Gini importance, permutation feature importance (PDP), Individual Conditional Expections (ICE), Lime, SHAP)

```{r explainability techniques}


```

# Prediction of our wages
Use model to predict our wages

```{r prediction of our wages}


```

