---
title: "Group Work 1 - Predictive modeling"
author: "Donjet Dzemaili, Fereshteh Ahmadi, Michael Etter, Seraina Zimmermann"
date: "2025-05-26"
output: html_document
---

# Introduction
What is the assignment all about. Donjet is the King! 

# Load and prepare the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(writexl)
library(readxl)

# Load the data
load("data_wage.RData")

# Rename for clarity
data_original <- data
rm(data)  # Optional: remove old name to avoid confusion

```

Make a copy from the original dataset to excel. so all of the group members can answer the questions and fill them directly in a separated row in the excel file. 
```{r export csv file to data}

# Export to Excel
write_xlsx(data_original, "data_wage_editable.xlsx")

```

After everyone filled the excel file out, we load the filled excel file. 

```{r upload filled excel file}

# Read the filled-in Excel file
data <- read_excel("data_wage_filled.xlsx")

```

```{r }
View(data)
names(data)
str(data)

```
# Feature Selection
Clear justification for selected features

# Data Quality Issues 

Handle missing values

Convert categorical variables to factors

(Optionally) one-hot encode factors

Normalize numeric features (standardisation)

```{r export csv file to data}

names(data)
str(data)

```

# Model choice
Explain the chosen model
Why did we chose this model


# Training
why 80/20 - explain

```{r export csv file to data}

names(data)
str(data)

```

# Testing
explain the evaluation metrics, why use them
Evaluate performance of model

# Explanation of the model
Run explainability techniques

```{r export csv file to data}

names(data)
str(data)

```

# Prediction of our wages

```{r export csv file to data}

names(data)
str(data)

```

